Writing Trash Bin

### 2024-02-21 paper.qmd
Finished the grammar of graphics section

*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline. Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Refsgaard2007]. @fig-location shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps. The relevance of every stage of the uncertainty taxonomy is backed up by both theoretical and practical work in statistics and adjacent fields. @Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore. @Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise. @Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input. High level stages such as model selection and estimation have obvious introductions of uncertainty through the choice of models and nature of estimation [^I'm not sure if I need a citation here because this feels obvious]. Location is also acknowledged as an important element of consideration in the taxonomies of several visualsiation papers [^ I will probably need a citation for which ones].

@@

@@

## Uncertainty Visualisation and the Grammar of Graphics
The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty.

- Since there are so many papers that treat uncertainty as a signal, not a noise, issues in the uncertainty visualisation literature also help to understand some difficulties in the information visualisation literature as large.

- Suggests that it is straightforward to show a value but much more complex to show uncertainty @Hullman2016 (I think this is interesting in relation to how people SEE uncertainty visualisation. Why is uncertainty hard to visualise?)
- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016
- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016
- Uncertainty defined as a by product of the process of making a visualisation, from data collection to commucating @Hullman2016
- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
- communication tasks: This category comprises map reading tasks involving data and uncertainty value retrieval (which location is most uncertain?). For tasks from this category, visualisation techniques can be chosen following the traditional rules from cartography. @Kinkeldey2014 (basically saying why dont you just use what is already here if you are treating uncertainty as a signal, which I agree with) 
- papers usually have same variance if there is a signal question, and use the mean as a proxy to answer the question. Really should ask a changing signal question where the signal IS impacted by the uncertainty information.
- Weirdly When setting their ground truth it seems like a lot of papers EXPECT participants to completely ignore uncertainty information as noise. Like the participants are actually considered to be incorrect if they dampen their signal understanding because of noise considerations which I find FUNNY. 
- I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.

Visual inference can be seen as a process that combines multiple fields, from mathematics to psychology, to convey meaning. Unfortunately which tasks belong to mathematics, computer science, or psychology is poorly defined. The uncertainty visualisation taxonomies make the blur between these fields. A visualisation is, in a lot of ways, a statistic, or at the very least statistics are calculated prior to the visualisation being calculated, however this may be invisible to the user. A density plot does not just "appear" from the data, usually a smoothing function (that should be calibrated according to the sample size however how this should be done is a different matter of dispute) generates an estimated density function (statistics) which is then depicted on a plot using a line (design or computer science) which is then converted into information in your head using perceptual tasks and heuristics (psychology).

These overlapping but distinct fields create a high burden of spanning expertise to understand if two visualisation are equivalent in each field. This is largely caused by the concept of "information" differing dramatically between fields. According to mathematics, a sufficient statistic for a parameter contains the same amount of information as an entire sample. Therefore, if you are trying to estimate a population mean, the sample mean and the entire data set are both equivalent, however a similar statistic, such as the median, is not and will become a worse and worse approximation for the mean the more skewed your underlying distribution in. Computer science considered two pieces of information to be equivalent if they come from the same data [^This is the justification that I have come across in several computer science field papers, I am not sure if this is true for the whole field so this is definitely a question for Sarah]. This definition of information is much looser and is contained within the mathematical definition of information.

Most uncertainty visualisation papers show that the definition of information *should* be the one defined by mathematics, not the one defined by computer science. A close look at the visual inference literature shows this to be true, but a large amount of noise is added to the research through visual heuristics. This is simply because, if one visualisation depicts a "significant statistic" and another visualisation shows a similar statistic, then the graphic that shows the significant statistic holds more relevant information to the question and will lead to more accurate interpretations. If the information that is depicted on a plot cannot be used to *mathematically* generate a best estimate for a specific statistic, I do wonder how the visual system would be expected to cover that gap if *not* for heuristics.

While mathematical information *should* be considered the starting point in considering if two visualisations are equivalent, it is not the only consideration. A large amount of work in perceptual tasks, attention, and psychology surrounding charts shows that all the information depicted is not paid attention to in equal measure. Elements such as colour can have sizable impacts on the way a graphic is percieved. Mathematics identifies the information that is *shown* while psychology identifies the difference in information *recieved*. A failure to understand this distinction leads to graphics that are different not only in the way information is depicted but in the information itself. This problem is pervasive in uncertainty visualisation literature.

- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.

- But when different techniques are to be compared, it is important that the scenarios and datasets are informationally equivalent, i.e. according to Larkin and Simon (1987, p. 67) ‘[t]wo representations are informationally equivalent if all of the information in the one is also inferable from the other, and vice versa’. A goal in testing, then, is often to determine whether they are also computationally equivalent, or whether one depiction has an advantage over another, i.e. ‘[t]wo representations are computationally equivalent if they are informationally equivalent and, in addition, any inference that can be drawn easily and quickly from the information given explicitly in the one can also be drawn easily and quickly from the information given explicitly in the other, and vice versa’ (Larkin and Simon 1987, p. 67). @Kinkeldey2014

- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016

- Most commonly check abolute measures of accuracy, where accuracy is abs(subjective error - actual probabiltiy). Some papers also check relative measures of accuracy and ask subjects to find regions of least certainty or to rank targets by uncertainty. studies also consider response time and confidence rating. some studies also ask users for their own expression of uncertainty (e.g. by asking when sketches are uncertainty and ranking various depictions of uncertainty). Some studies ask subjects to make some decision using the data. ometimes things like complexity of task, degree of visual overload, ease of use, visual appeal, and preferences @Hullman2016

## Noise and Signal in Uncertainty Visualisation
The first issue that arises with uncertainty research is that uncertainty is frequently treated as a signal, not noise. This discrepancy is also noted by @Kinkeldey2014 who, in their own literature review on uncertainty methods, comment that the literature seems unsure if uncertainty is simply another variable or if it should be treated differently. This is a fair question to ask. If uncertainty is simply a signal we are trying to convey, why should it be treated any different to any other signal in a visualisation?

@@

## An Unexplored Space (Different visualisations for location and stuff)
- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty). Other tasks include comparisons, rankings. also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014. e.g. land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates). If you combine them all it might not make sense for someone who only needs one.

## Contextual Inormation
I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty. I did not go into too much detail because it would involve discussing visualisation which I am trying to avoid at this point. I actually think this quote from the grammar of graphics is exactly what I was trying to say. *"The system in this book is not a model for a GIS, however, because geography and statistics differ in a crucial respect. Geography is anchored in real space-time and statistics in abstract dimensions. This is a distinction along a continuum rather than a sharp break; after all, there is a whole field called spatial statistics (Cressie, 1991). But this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics"*. Basically, my point is this section is still discussing the definitions of unertainty, we have not gotten to disussing visualisation so we are still in the abstract *statistics* domain. The only parts of spatial statistics that we care about in this domain are the statistical properties that come from spatial information, and *none* of those statistical properties generate uncertainty. This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and missing information generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data. *This does not mean I am saying spatial and temporal information is not valuable ever*, I am just saying it is not valuable *yet*. This information will be *highly relevant* when we discuss context and translating from the abstract domain back into real world for communication. Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature. The confusion that is created in the literature was going to be the topic I had later. Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised, not what they should *try* to visualise based on the uncertainty abstraction. Since the context is only relevant *with respect to uncertainty* when we visualise it, I was going to discuss this issue in the visualisation literature review section.

>>>>

*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline. Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @fig-location shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps. The relevance of every stage of the uncertainty taxonomy is backed up by both theoretical and practical work in statistics and adjacent fields. @Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore. @Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise. @Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input. High level stages such as model selection and estimation have obvious introductions of uncertainty through the choice of models and nature of estimation [^I'm not sure if I need a citation here because this feels obvious]. Location is also acknowledged as an important element of consideration in the taxonomies of several visualsiation papers [^ I will probably need a citation for which ones].

@@

::: aside
I should maybe change this to relatability to uncertainty, value extraction, decision making and trust, and meta papers or something. Maybe perception and evaluation papers?
:::

@@

## Perception Experiments and a Poor Definition
Even the most simple experiments are not free from the issues in the uncertainty definition.

Papers that consider unique aspects of uncertainty in their experiment such as experiemnts that identify the usefulness of specific perceptual tasks, experiments that identify problematic heuristics in commonly used uncertainty visualisations, and experiments that consider unique problems in uncertainty experiments all have useful results. These papers will be discussed in the upcomming sections.

Perceptual task experiments that identify visual aesthetics that are intuitively connected to uncertainty are useful as they establish an inutive psychological connection. These papers are also not without their flaws, however their flaws have little to do with issues in the grammar of graphics as they are well defined in this space and more to do with the lack. Since uncertainty and signal are typically mapped to different aesthetics, it is not always clear what exactly the visualisation is suggesting we should be uncertain about, and when a perceptual task that should be considered something like a train line has "uncertainty" added to it, it is not

- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty). Other tasks include comparisons, rankings. also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014. e.g. land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates). If you combine them all it might not make sense for someone who only needs one.

### Contextual Inormation
I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty. I did not go into too much detail because it would involve discussing visualisation which I am trying to avoid at this point. I actually think this quote from the grammar of graphics is exactly what I was trying to say. *"The system in this book is not a model for a GIS, however, because geography and statistics differ in a crucial respect. Geography is anchored in real space-time and statistics in abstract dimensions. This is a distinction along a continuum rather than a sharp break; after all, there is a whole field called spatial statistics (Cressie, 1991). But this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics"*. Basically, my point is this section is still discussing the definitions of unertainty, we have not gotten to disussing visualisation so we are still in the abstract *statistics* domain. The only parts of spatial statistics that we care about in this domain are the statistical properties that come from spatial information, and *none* of those statistical properties generate uncertainty. This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and missing information generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data. *This does not mean I am saying spatial and temporal information is not valuable ever*, I am just saying it is not valuable *yet*. This information will be *highly relevant* when we discuss context and translating from the abstract domain back into real world for communication. Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature. The confusion that is created in the literature was going to be the topic I had later. Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised, not what they should *try* to visualise based on the uncertainty abstraction. Since the context is only relevant *with respect to uncertainty* when we visualise it, I was going to discuss this issue in the visualisation literature review section.

## Evaluation Experiments and The Grammar of Graphics
The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty. This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments. How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies? The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.

::: aside
I think I want the "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point here since this section is now about value extraction.
:::

Visual inference can be seen as a process that combines multiple fields, from mathematics to psychology, to convey meaning. Unfortunately which tasks belong to mathematics, computer science, or psychology is poorly defined. The uncertainty visualisation taxonomies make the blur between these fields. A visualisation is, in a lot of ways, a statistic, or at the very least statistics are calculated prior to the visualisation being calculated, however this may be invisible to the user. A density plot does not just "appear" from the data, usually a smoothing function generates an estimated density function, which is then depicted on a plot using a line, which is then converted into information in your head using perceptual tasks and heuristics.  These overlapping but distinct fields create a high burden of spanning expertise to understand if two visualisation are equivalent in each field. This is largely caused by the concept of "information" differing dramatically between fields. According to mathematics, a sufficient statistic for a parameter contains the same amount of information as an entire sample. Therefore, if you are trying to estimate a population mean, the sample mean and the entire data set are both equivalent, however a similar statistic, such as the median, is not and will become a worse and worse approximation for the mean the more skewed your underlying distribution in. Computer science considered two pieces of information to be equivalent if they come from the same data. This definition of information is much looser and is contained within the mathematical definition of information. Thanksfully, we can turn to the grammar of graphics to try and organise and untangle the information in our graphics.

When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not. @Leland2005 depicts these steps as a pipeline, shown in @fig-pipeline. This pipeline shows every step from data observation to rendering of a graphic, but we are going to focus on the tail end of the pipeline specifically on statistics, geometry, coordinates, and aesthetics.

![The grammar of graphics data analysis pipeline](grammarofgraphicspipeline.png){#fig-pipeline}

In order to visually communicate uncertainty we must:  
1) Mathematically quantify uncertainty in some way. 
2) Visually express this quantification using some geometry, axis, aesthetic, and rendering.
3) Interpret the visual expression to extract the information using perceptual tasks.

Each of these steps have a large amount of research independent of the "uncertainty visualisation" subfield. There are numerous ways to depict uncertainty mathematically. Examples of mathematically defined objects that are used to express uncertainty inlude: confidence intervals, sampling distributions, bias, variance, prediction error, value ranges, quantiles etc. These definitions all exist in mathematics independent of the field of visualisation. Once we have some quantification of uncertainty, we need to express it visually. There are numerous options in combinations of geometry, axis, and aesthetics and modern graphical software will allow most of these combinations. There is also a large amount of research in information visualisation and psychology into how the choice of visual expression impacts a persons ability or ease with which they extract relevant information. In this sense, once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.

This distinction is important because lower levels of the pipeline influence higher levels. Visualisation authors are almost unanimous in commenting that the information in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] *this citation has way more, check confirmation*. @Kinkeldey2014 adopts a definition from the 80s that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice versa, and computationally equivalent if that information can be extracted from both plots with similar easy and speed. This definition illustrates a similar idea to the grammar of graphics pipeline, shown in @fig-pipeline. First the mathematical information in a plot is established, and then that mathematical information is visually represented. Two plots that depict mathematical information that cannot be inferred from each other, cannot be compared on their visual elements. To prevent this requirement from being *too* limiting, when using a ground truth statistic we can consider two plots to be informationally equivalent if both visualisations depict a sufficient statistic. For example, a depiction of a sample and its average are equivalent if we are trying to infer the population mean. This does not mean they are computationally equivalent. Additionally, two plots cannot be compared on their visual elements if they contain different information as well. The reason for this is obvious. If you compare two graphics and one depicts the mean and the other depicts a median and you ask viewers to extract the median value, the graphic that *actually* depicts that statistic will perform better. People cannot percieve information that is not depicted.

Computational equivalency in visualisation is also something that has been well researched. For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others. The hierarchy established 40 years ago by @Cleveland1984 is: 
1) position along a common scale
2) position along a non-aligned scale
3) length
4) angle/slope
5) area
6) volume
7) colour
This hierarchy has had additions and adjustments over the years. For example, this hierarchy is a general rule, but the hierarchy can change from individual to individual. Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles. These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.

A bizarre feature of uncertainty visualisation literature is that it does not work to build upon these existing principle or identify the ways in which uncertainty visualisations may diverge from these rules, but rather it seems to exist with complete indifference to the research in information visualisation at large. The most common uncertainty visualisation evaluation study asks participants to perform some value abstraction either explicitly by asking participants to report values or vaguely by asking participants to find regions of least certainty, rank plots, report their confidence, and/or make decisions according to uncertainty information. Experimenters then compare these results based on the participants accuracy, preference, confidence, and/or speed @Hullman2016. Most commonly, these value abstraction papers compare two different visualisations of uncertainty on a set of these questions, however, since there is no rules for what is or is not an uncertainty visualisation the graphics that are compared often differ in both computation *and* information. Uncertainty visualisaiton papers only seem to require that the plots depict "uncertainty" which has already been established to be poorly defined. The complete lack of formality is easy to see if we actually plot the visualisations compared across evaluation experiments according to the grammar of graphics *in the figure below*:

**INCLUDE A GRAPHIC THAT SHOWS WHERE EACH TYPE OF UNCERTAINTY VISUALISATION SITS IN THE GRAMMAR OF GRAPHICS AND HOW DIFFERENT THEY ARE ACROSS MULTIPLE STUDIES**

Reading the uncertainty visalisation literature will make the importance of information and computation incredibly salient. If plots that have different information are compared, whichever directly depicts the ground truth statistic (or has information that can easily be convered to that ground truth statistic) will perform the best *cite*. If plots that have the same information are compared, the plot that requires a perceptual task that is higher on the hierarchy will perform the best *cite*. Therefore, when it comes to any task that involves value abstraction, many uncertainty visualisation evaluation studies are simple repeating already established results in visualisation.

Several uncertainty visualisation literature reviews have commented on the immense difficulty of combining results across experiments, but none have been able to pinpoint the reason. @Kinkeldey2014 suggested a task based taxonomy that organised the plots by the question or purpose of the visualisation, and even suggested that communication tasks (such as value retrival of uncertainty values) should just follow traditional cartography rules, @Spiegelhalter2017 simply ignored the existence of uncertainty visualisation as a subfield and just suggested we follow established rules within uncertainty visualisation, @Hullman2016 said that commented on the large amount of noise generated from uncertainty visualisation evaluation methods that led to a difficulty in combining results, @Griethe2006 did not meaningfully combine the work and instead listed findings in the field, but also did not comment on the lack of generalisable results. While we largely agree with the difficulties of these authors, we do not agree that the solution is a task based taxonomy or better questions. While we agree the lack of attention to tasks and appropriate questions has generated problems in the field, there is a much more pressing issue, which is that uncertainty visualisation authors don't seem to understand *why* uncertainty visualisation different to typical visualisations.

The class of "uncertainty visualisation" does not seem to have any rules for what belongs in the group and what makes it different to a "typical" visualisation is rarely discussed. Instead one of two general motivations for the field are provided in the subtext.  
1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics.  
Neither of these motivations are remotely relevant when we are evaluating the simple task of value extraction. The psychological effects of uncertainty are only relevant for decision making and have nothing to do with value abstraction, and the information overload issue is not specific to uncertainty and is simply an example of a larger dimensionality issue. Therefore, these explanations for the difference fall short of explaining *why* uncertainty should be treated any differently than any other variable.

Despite these results, it is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been apparent in existing uncertainty visualisation experiments. @Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty. So, if we cannot see any difference to uncertainty information when it coems to value abstraction, *when* is uncertainty information different?

## Noise and Signal in Uncertainty Visualisation
Uncertainty is noise, not signal. While this comment may seem obvious, what may not be obvious is that the value abstraction experiments are treating uncertainty as *signal*. That is why the experimental results are identical to the existing information visualisation literature.

This discrepancy is also noted by @Kinkeldey2014 who, in their own literature review on uncertainty methods, comment that the literature seems unsure if uncertainty is simply another variable or if it should be treated differently. This is a fair question to ask. If uncertainty is simply a signal we are trying to convey, why should it be treated any different to any other signal in a visualisation?

- I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.
- Weirdly When setting their ground truth it seems like a lot of papers EXPECT participants to completely ignore uncertainty information as noise. Like the participants are actually considered to be incorrect if they dampen their signal understanding because of noise considerations which I find FUNNY. (this feels like it is more as a communicating uncertainty issues)
- papers usually have same variance if there is a signal question, and use the mean as a proxy to answer the question. Really should ask a changing signal question where the signal IS impacted by the uncertainty information.
- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016 (moved here because it is clear she is touching on the whole "what are we uncertain about" thing)

@@

### 2024-02-21 writingdumpster.qmd
lot of additions to the grammar of graphics section
Finished the grammar of graphics section

# Moved over notes
- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
-  @Spiegelhalter2017 defines risk as risk known probabilities and uncertainty as estimated probabilities, therefore uncertainty is what we have when you cannot accurately quantify risk.

>>>>

# Moved over notes
- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
-  @Spiegelhalter2017 defines risk as risk known probabilities and uncertainty as estimated probabilities, therefore uncertainty is what we have when you cannot accurately quantify risk.
- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016 
- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016

# Main section outtakes
While mathematical information *should* be considered the starting point in considering if two visualisations are equivalent, it is not the only consideration. A large amount of work in perceptual tasks, attention, and psychology surrounding charts shows that all the information depicted is not paid attention to in equal measure. Elements such as colour can have sizable impacts on the way a graphic is percieved. Mathematics identifies the information that is *shown* while psychology identifies the difference in information *recieved*. A failure to understand this distinction leads to graphics that are different not only in the way information is depicted but in the information itself. This problem is pervasive in uncertainty visualisation literature.

This is not to say any paper that discusses uncertainty visualisation is completely pointless because the results already exist. We are simply pointing out that treating uncertainty graphics as special simply because they contain uncertainty, even if that uncertainty is not related to the task, leads to results that are already established in information visualisation.

These two perceptualisations of *why* uncertainty visualisation is different from normal visualisation is also what leads to differing opinions on what is an uncertainty visualisation. For example, someone who percieves of uncertainty visualisation as a dimensionality problem would see a pie chart as a normal graphic, and a pie chart with blurred boundaries as an uncertainty visualisation (even though blurring the boundary between segments of a pie chart makes no sense when you think of a pie chart as a polar co-ordinate bar chart), while someone who sees uncertainty visualisation as psychologically different would see a pie chart as an uncertainty graphic so long as it depicts probabilities. This may lead you to believe that those who believe uncertainty is unique due to its psychological properties believe uncertainty should be defined at the mathematical stage, while those that believe it is an addition to an established visualisation think uncertainty is defined in a graphic at the aesthetic stage. This is not the case. As shown by the *grammar of graphics fig* there is no single person or experiment that is consistent in any one belief. This difficulty and lack of consistency comes from the desire to discuss "uncertainty visualisation" as though it is a single term and not the combination of "uncertainty" and "visualisation". If we split the terms and look at the quantification and visualisation of uncertainty separately, the results of many uncertainty visualisation papers become self evident.

Papers that are justified using the first type of motivation rarely explain exactly how the psychological apsects of uncertainty impct the participants in a way that inhibits their ability to read information from a graphic. Since uncertainty is poorly defined and the psychological effects of uncertainty is never justifiably connected to the scientific communication of it, as was discussed in ssections 3 and 4, it is unclear exactly *what* these papers are testing for. It is almost as though they *feel* that uncertainty visualisations *should* be different, but do not consider *how* uncertainty information is supposedly different beyond decision making criteria such as *risk avoidance* which is not something science communicators have any business trying to control, as was discussed in previous sections. For these reasons, experiments that test for secondary factors (such as influence on decision making) are confounding far too many variables to establish any generalisiable take away, and experiments that ask participants to extract values fail to notice there is no longer a reason for "uncertainty" to be any different to any other type of visualisation.

Papers that are justified using the second type of motivation clearly fit within the sub-field of high dimensional visualisation, but also rarely comment on the connection. These experiments evaluate different ways to include even *more* information than would be put in a typical visualisation. Some of these methods, such as a bivariate pallet or glyph map, create graphics that are overloaded with information and offer incorrect or meaningless takeaways at first glance. This is not to say *every* visualisation method that sees it as a dimensionality problem is bad. Clever methods of combining several types of information, such as value supressing uncertainty pallets (VSUPs), can better help us understand how our brains process information in a graphic.  For example, the VSUP allows us to test is colour value and saturation can be combined such that we mentally combine signal and noise together as one variable. Since the take aways from the papers are not usually related to uncertainty specifically, and often related to interesting perceptual tasks, there is also no justification for why this work is "uncertainty visualisation" work and not simply perceptual task studies that can be used by the rest of the information visualisation community.

This stems from a class of "uncertainty visualisation" that seems to have no rules for what belongs in the group, and no consideration for the mathematics that is actually depicted in these plots. Uncertainty visualisation experiments do not work to build upon existing visualisation literature or differentiate uncertainty visualisaiton as a new field, but seem to exist with shocking indifference to the field it belongs to.

uncertainty visualisation, as it is now, should not exist. Any generalisable contribution made by these papers is already an established result in information visualisaiton, and therefore is not specific to "uncertainty visualisation".  Any "uncertainty visualisation" specific result is not generalisable beyond the experimental conditions because of the poor formalisation in the definitions of uncertainty and uncertainty visualisations.

This lack of formalisation is not unique to uncertainty visualisation and the wider field of information visualisation suffers form similar problems [@Kinkeldey2014].

This makes it clear we have a serious problem with the classification of "uncertainty visualisation". The visualisations follow identical rules to those established already in the visualisation literature, and experiments that view uncertianty visualisation as somehow special or different from normal visualisation are simply repeating existing established research in information visualisation.

If several graphics are compared that are different statistically, geometrically, and aesthetically, there is *no way to combine the results to establish generic principles for the field*.

The overwhelming majority of these papers depict different mathematical expressions of both the signal and the uncertainty. In 100% of these papers, the plot that has mapped the statistic in question to some visual feature will outperform those that don't. When the statistics used to depict uncertainty *are* the same (which is rare) 100% of those papers will follow the hierarchy of perceptual tasks that has already been established in the visualisation literature. If the experiment participant needs to extract a probability and one graphic uses position along a common scale, while the other uses colour or literally any other perceptual task, the position depiction will always outperform the other.

If we were to compare two different visualisations, we would like to change as few features as possible so we can better understand the cause of the difference in perception.

Additionally, if information that can be used to calculate a statistic, but not the statistic itself, is shown, people will need to use a heuristic to extract that information which will decrease the precision with which that information is communicated. If the information that is depicted on a plot cannot be used to *mathematically* generate a best estimate for a specific statistic, it is questionable how uncertainty visualisation authors expect participants to cover that gap if *not* for heuristics.

In these papers will typically ask separate questions about the signal (e.g. what is the mean of this distribution) and the uncertainty (e.g. what is the variance of this distribution).

Failling to understand this key aspect of visualsiation leads to graphics that *contain* the correct amount of information, but that volume of information cannot be readily accessed by the viewer.
